A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-15 text/informal-overview.tex

We designed \LMD, a multi-stage calculus with dependent type.
\LMD is based on \LTP\cite{Hanada2014} by Hanada and Igarashi, which is a multi-stage calculus with CSP and
we introduced dependent types based on \LTP\cite{attapl}.
In this section, we check \LMD informally after checking \LTP and \LLF which are the basis of \LMD.

\red{\LTP\cite{Hanada2014}が頻発するが同一論文の複数回引用に何かルールはあるのか?}
\subsection{\LTP}

@@

The subscript $\alpha$ in $\TB_\alpha M$ is a \textit{transition variable} and
a sequence of transition variables is called \textit{transition}.
The empty transition is represented by $\varepsilon$.
They are used to show the thickness of brackets.
For example, $\TB_\alpha (\lambda x:\I.x+10)$ is the fragment of code which becomes $(\lambda x:\I.x+10)$ after it was run once and
$\TB_\alpha \TB_\beta (\lambda x:\I.x+10)$, which is abbreviated as $\TB_{\alpha\beta} (\lambda x:\I.x+10)$,
becomes $(\lambda x:\I.x+10)$ after it was run it twice.
\red{run twiceは表現が微妙}

@@

There are abstractions for transition variables and applications for transition abstraction in \LTP.
They look like $\Lambda\alpha.M$ and $M A$, respectively.
A transition abstraction binds a transition variable in a term.
For example, all $\alpha$ in $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))$ are bound.
It is only natural there is a reduction rule for transition application in \LTP.
The rule is the following.
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
For example, $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))\ (\beta\gamma)$ reduces to $\TB_{\beta\gamma} (\lambda x:\I.x)$.

@@

CSP, cross-stage persistence, is an important feature of \LTP.
It enables us to embed value at an outer transition into an inner transition.
$\%$ is dedicated to CSP in \LTP.
For example, $\lambda a:\I.\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x+\%_\alpha a))$
\red{inner / outer は適切か?}

@@

% \LLF
\LLF is a simple system of dependent types introduced in \cite{attapl}.
\red{system / type system / calculus?}
It is based on Edinburgh LF\cite{harper1993framework}.
Therefore, all constants and base types are declared in the signature.
The \LLF type theory generalizes simply typed lambda calculus
by replacing the function type $\tau\to\sigma$ with the dependent product type $\Pi x:\tau.\sigma$.

% Kind, Well-formed kind
In addition to ordinary typing rules like simply typed lambda calculus,
there are kinding rules, well-formed kinding rules, term equivalence rules, type equivalence rules, and kind equivalence rules in \LLF.
Kinding rules and well-formed kinding rules are 
introduced in order to prohibit making illegal types such as $\textrm{Vect}\ \textrm{Bool}$.
For a well-formed type $\tau$, $\G \vdash \tau :: K$ means that $\tau$ has a kind $K$ under the environment $\G$ and 
for a well-formed kind $K$, $\G \vdash K$ means that $K$ is a well-formed kind under an environment $\G$.

@@

Third, we also need type equivalence rules in \LMD because its type system contains dependent types.
Although there are new primitives relating to stages which aren't in \LLF,
we can design all rules except \QPercent\ easily from the syntax or the reductions of \LMD.

\QPercent\ is different from other rules because it was designed from practical reason, not from the syntax or the reductions.
For example, we discuss $\text{vadd}_\beta$ function,
which takes a length of vector and returns addition function for vectors with the length.
$\text{vadd}_1$ is a helper function to make $\text{vadd}$, 
which takes a stage, a length of vectors, and two quoted vectors and returns a quoted vector.
We assumed that type checker can know $n=0$ when process the $\textbf{then}$ clause.
\newcommand{\Vpn}{\text{Vector}\ (\%_\alpha n)}
\begin{align*}
	  & \textbf{let}\ \text{vadd}_1 : \F\alpha.\Pi n:\I.\TW_\alpha\Vpn\to\TW_\alpha\Vpn\to\TW_\alpha\Vpn                                \\
	  & \hspace{7mm} = \textbf{fix}\ f.\Lambda\alpha.\lambda n:\I.\ \lambda v_1:\TW_\alpha\Vpn.\ \lambda v_2:\TW_\alpha\Vpn.            \\
	  & \hspace{14mm} \textbf{if}\ n = 0                                                                                                \\
	  & \hspace{14mm} \textbf{then} \TB_\alpha \text{nil}                                                                               \\
	  & \hspace{14mm} \textbf{else}\ \TB_\alpha (                                                                                       \\
	  & \hspace{21mm} \textbf{let}\ t_1 = \text{tail}\ (\TBL_\alpha v_1)\ \textbf{in}                                                   \\
	  & \hspace{21mm} \textbf{let}\ t_2 = \text{tail}\ (\TBL_\alpha v_2)\ \textbf{in}                                                   \\
	  & \hspace{21mm} \text{cons}\ (\text{head}\ (\TBL_\alpha v_1) + \text{head}\ (\TBL_\alpha v_2))\ (\TBL_\alpha f\ (n-1)\ t_1\ t_2)) \\
\end{align*}
Using $\text{vadd}_1$, we can make $\text{vadd}$ function easily.
This function takes the length of vectors and returns a quoted add function for vector of the given size.
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta n)}
\begin{align*}
	  & \textbf{let}\ \text{vadd}: \Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)                \\ 
	  & \hspace{7mm} = \lambda n:\I.\Lambda\beta.\TB_\beta (\lambda v_1:\Vpn.\ \lambda v_2:\Vpn. \\
	  & \hspace{63mm} \TBL_\beta \text{vadd}_1\ \beta\ n\ (\TB_\beta\ v_1)\ (\TB_\beta\ v_2))    \\
\end{align*}
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta 10)}
When we use this function, we need to instantiate $\text{vadd}$ with a length.
If we instantiate it with a integer 10, we get function $\text{vadd10}$ whose type is \\
$\text{vadd10}: \F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)$.
This function is difficult to use because its type contains $\Vpn$.
In most cases, we have vectors of type $\text{Vector}\ 10$, not $\Vpn$.
Therefore, we want to identify $\text{Vector}\ 10$ and $\Vpn$.

>>>>

We designed \LMD, a multi-stage calculus with dependent type.
\LMD is based on \LTP\cite{Hanada2014} by Hanada and Igarashi, which is a multi-stage calculus with CSP, and
we introduced dependent types based on \LTP\cite{attapl}.
In this section, we check \LMD informally after checking \LTP and \LLF which are the basis of \LMD.

\red{\LTP\cite{Hanada2014}が頻発するが同一論文の複数回引用に何かルールはあるのか?}
\AI{ないです．たしかにあんまりつけるとうるさいですね．2.1 の冒頭とかは文ごと不要なのでは？}
\subsection{\LTP}

@@

The subscript $\alpha$ in $\TB_\alpha M$ is a \textit{transition variable} and
a sequence of transition variables is called \textit{transition}.
The empty transition is represented by $\varepsilon$.
They are used to show the thickness of brackets. \AI{Perhaps ``thickness'' doesn't make sense.}
For example, $\TB_\alpha (\lambda x:\I.x+10)$ is the fragment of code which becomes $(\lambda x:\I.x+10)$ after it was run once and
$\TB_\alpha \TB_\beta (\lambda x:\I.x+10)$, which is abbreviated as $\TB_{\alpha\beta} (\lambda x:\I.x+10)$,
becomes $(\lambda x:\I.x+10)$ after it was run it twice.
\red{run twiceは表現が微妙}

@@

There are abstractions for transition variables and applications for transition abstraction in \LTP.
They look like $\Lambda\alpha.M$ and $M A$, respectively.
A transition abstraction binds a transition variable in a term.
For example, all $\alpha$ in $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))$ are bound.
It is only natural \AI{?} there is a reduction rule for transition application in \LTP.
The rule is the following.
\begin{align*}
	(\Lambda\alpha.M)\ A \longrightarrow M[\alpha\mapsto A]
\end{align*}
For example, $\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x))\ (\beta\gamma)$ reduces to $\TB_{\beta\gamma} (\lambda x:\I.x)$.

@@

CSP, cross-stage persistence, is an important feature of \LTP.
It enables us to embed value at an outer transition into an inner transition.
$\%$ is dedicated to CSP in \LTP.
\AI{What should me mentioned here is that in MetaOCaml CSP is implicit but
in \LTP it's explicit.}
For example, $\lambda a:\I.\Lambda\alpha.(\TB_\alpha (\lambda x:\I.x+\%_\alpha a))$
\red{inner / outer は適切か?}
\AI{What is \(a\)?}
\AI{Is this explanation enough to understand the rest of the section?}

@@

% \LLF
\LLF is a simple system of dependent types introduced in \cite{attapl}.
\red{system / type system / calculus?}
It is based on Edinburgh LF\cite{harper1993framework}.
\AI{Can you remind me of differences between Edinburgh LF and \LLF?}
Therefore, \AI{``Therefore'' is used to show a logical consequence.} all constants and base types are declared in the signature.
The \LLF type theory generalizes simply typed lambda calculus
by replacing the function type $\tau\to\sigma$ with the dependent product type $\Pi x:\tau.\sigma$.  \AI{``dependent product'' vs. ``dependent function''.}

% Kind, Well-formed kind
In addition to ordinary typing rules like simply typed lambda calculus,
there are kinding rules, well-formed kinding rules, term equivalence rules, type equivalence rules, and kind equivalence rules in \LLF.
Kinding rules and well-formed kinding rules are 
introduced in order to prohibit making illegal types such as $\textrm{Vect}\ \textrm{Bool}$.  \AI{Vect\ true may be a better example?}
For a well-formed type $\tau$, $\G \vdash \tau :: K$ means that $\tau$ has a kind $K$ under the environment $\G$ and 
for a well-formed kind $K$, $\G \vdash K$ means that $K$ is a well-formed kind under an environment $\G$.

@@

Third, we also need type equivalence rules in \LMD because its type system contains dependent types.
Although there are new primitives relating to stages which aren't in \LLF,
we can design all rules except \QPercent\ easily from the syntax or the reductions of \LMD.  \AI{Don't refer to a concrete rule name without showing it because we haven't introduced rules.}

\AI{If there's space, we should have introduced the \LTP-version of vadd.}

\QPercent\ is different from other rules because it was designed from practical reason, not from the syntax or the reductions.
For example, we discuss $\text{vadd}_\beta$ function,
which takes a length of vector and returns addition function for vectors with the length.
$\text{vadd}_1$ is a helper function to make $\text{vadd}$, 
which takes a stage, a length of vectors, and two quoted vectors and returns a quoted vector.
We assumed that type checker can know $n=0$ when process the $\textbf{then}$ clause.
\newcommand{\Vpn}{\text{Vector}\ (\%_\alpha n)}
\begin{align*}
	  & \textbf{let}\ \text{vadd}_1 : \F\alpha.\Pi n:\I.\TW_\alpha\Vpn\to\TW_\alpha\Vpn\to\TW_\alpha\Vpn                                \\
	  & \hspace{7mm} = \textbf{fix}\ f.\Lambda\alpha.\lambda n:\I.\ \lambda v_1:\TW_\alpha\Vpn.\ \lambda v_2:\TW_\alpha\Vpn.            \\
	  & \hspace{14mm} \textbf{if}\ n = 0                                                                                                \\
	  & \hspace{14mm} \textbf{then} \TB_\alpha \text{nil}                                                                               \\
	  & \hspace{14mm} \textbf{else}\ \TB_\alpha (                                                                                       \\
	  & \hspace{21mm} \textbf{let}\ t_1 = \text{tail}\ (\TBL_\alpha v_1)\ \textbf{in}                                                   \\
	  & \hspace{21mm} \textbf{let}\ t_2 = \text{tail}\ (\TBL_\alpha v_2)\ \textbf{in}                                                   \\
	  & \hspace{21mm} \text{cons}\ (\text{head}\ (\TBL_\alpha v_1) + \text{head}\ (\TBL_\alpha v_2))\ (\TBL_\alpha f\ (n-1)\ t_1\ t_2)) \\
\end{align*}
\AI{I think we should state the types of cons, tail, and head.}
Using $\text{vadd}_1$, we can make $\text{vadd}$ function easily.
This function takes the length of vectors and returns a quoted add function for vector of the given size.
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta n)}
\begin{align*}
	  & \textbf{let}\ \text{vadd}: \Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)                \\ 
	  & \hspace{7mm} = \lambda n:\I.\Lambda\beta.\TB_\beta (\lambda v_1:\Vpn.\ \lambda v_2:\Vpn. \\
	  & \hspace{63mm} \TBL_\beta \text{vadd}_1\ \beta\ n\ (\TB_\beta\ v_1)\ (\TB_\beta\ v_2))    \\
\end{align*}
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta 10)}
When we use this function, we need to instantiate $\text{vadd}$ with a length.
If we instantiate it with a integer 10, we get function $\text{vadd10}$ whose type is \\
$\text{vadd10}: \F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)$.
This function is difficult to use because its type contains $\Vpn$.
In most cases, we have vectors of type $\text{Vector}\ 10$, not $\Vpn$.
Therefore, we want to identify $\text{Vector}\ 10$ and $\Vpn$.

### 2019-06-15 text/introduction.tex
分割コンパイルが失敗する

Multi-stage programming enables us to generate and run fragments of code at runtime.
MetaOCaml\cite{oleg2014} provide features for multi-stage programming, which are brackets, escape, and run.
Brackets, written \verb|.< e >.| in MetaOCaml, makes a code value from \verb|e|.
Run, written \verb| run e | in MetaOCaml, runs a code value of \verb|e| and restore \verb|e|.
Escape, written \verb| .~ e | in MetaOCaml, expands a code value of \verb|e|.
Unlike run, escape is supposed to use in a fragment of code, that is, escape cannot be used as an alternative of run.
For example, the following MetaOCaml expression

@@

Cross-stage Persistence (CSP) is another primitive of multi-stage programming, which enable us to embed computed values into a code value.
This is an example of CSP.
\begin{verbatim}
        let plusone = fun x -> x + 1 in
        let y = plusone 2 in  .< y >.
\end{verbatim}
This program evaluates to \verb|.< 3 >.| not  \verb|.< plusone 2 >.|.
This is because the variable \verb|y| is introduced into a code fragment using CSP.
Therefore, the variable \verb|y| is embedded after it was calculated.
CSP is very important when we write practical programs 
because CSP enables us to use library functions in code fragments as in \verb|.<List.combine [1;2] ['a';'b']>.|.

@@

However, there is a sever problem in functions which are optimized by multi-stage programming.
Although unrolled \verb|vadd| function is optimized, we cannot use it for different vector length.
For example, when you optimize for the length of 5, you shouldn't use it for vectors of 3 lengths.
Otherwise, we will get a Segmentation Fault error.
This problem is serious but existing type systems for multi-stage calculi cannot prevent it.

@@

@@

Dependent types are types which is dependent on values.
We can use dependent types for securer programming.
For example, we can realize vectors with their sizes with them.
\begin{verbatim}
        Vector :: Int -> *
\end{verbatim}
\verb|Vector| is a type constor which takes the length of vectors, that is, 
\verb|Vector 3| is a type for vectors whose lengths are 3.
If \verb|vadd| function has the type of \verb|Vector n -> Vector n -> Vector n|,
we can confirm the two arguments of \verb|vadd| function has the same length.
\red{secure programming? finer grained typing? とか}

\subsection{Multi-stage Programming with Dependent Types}
\red{このsubsectionは短いので上とくっつけてもいいかもしれない}

% \subsubsection{「生成した行列計算のコードは特定のサイズに特化しており、他のサイズでは使えない。
% このため、間違って組み合わせると容易にエラーを引き起こす。」問題を依存型で解決する。}

@@

In this paper, we design new multi-stage calculus \LMD by 
merging dependent types into existing multi-stage calculus \LTP\cite{Hanada2014}.
\LMD is multi-stage calculus which contains dependent types.
\LMD makes multi-stage programming more safer with the power of dependent types.

@@

The organization of this paper is the following.
Section 2 gives an informal overview of \LMD.
Section 3 defines the syntax, type system, full reduction, evaluation contexts of \LMD.
Section 4 shows the properties of \LMD including Unique Decomposition.
Section 5 discusses future works.
Finally, Section 6 discusses related works.
\AI{Use $\backslash$label and $\backslash$ref!}
\AI{``Related work'' and ``future work'' (they are uncountable).}

>>>>

Multi-stage programming enables us to generate and run fragments of code at runtime.  \AI{It sounds like the subject of ``generate and run'' is human being (``us'').  Something like ``Multi-stage programming offers linguistic support for dynamic code generation and execution.''?}
MetaOCaml~\cite{oleg2014} \AI{You need a space between a word and a citation.  Taha's original MetaOCaml should be cited, too.} provides features for multi-stage programming, which are brackets, escape, and run.  \AI{CSP?}
Brackets, written \verb|.< e >.| in MetaOCaml, makes a code value from \verb|e|.
Run, written \verb| run e | in MetaOCaml, runs a code value of \verb|e| and restore \verb|e|.
Escape, written \verb| .~ e | in MetaOCaml, expands a code value of \verb|e|.
Unlike run, escape is supposed to use in a fragment of code, that is, escape cannot be used as an alternative of run.
For example, the following MetaOCaml expression

@@

Cross-stage persistence (CSP) is another primitive of multi-stage programming, which enable us to embed computed values into a code value.
This is an example of CSP.
\begin{verbatim}
        let plusone = fun x -> x + 1 in
        let y = plusone 2 in  .< y >.
\end{verbatim}
This program evaluates to \verb|.< 3 >.| not  \verb|.< plusone 2 >.|.
This is because the variable \verb|y| is introduced into a code fragment using CSP.
Therefore, the variable \verb|y| is embedded after it was calculated.
CSP is very important when we write practical programs 
because CSP enables us to use library functions in code fragments as in \verb|.<List.combine [1;2] ['a';'b']>.|.

@@

However, there is a sever problem in functions which are optimized by multi-stage programming.
Although unrolled \verb|vadd| function is optimized, we cannot use it for different vector length.
For example, when you optimize for the length of 5, you shouldn't use it for vectors of 3 lengths.
Otherwise, we will get a Segmentation Fault error.
\AI{Why SEGV?  MetaOCaml will raise an exception.}
This problem is serious but existing type systems for multi-stage calculi cannot prevent it.

@@

\AI{このイントロ、各段落はまあまあよいのですが、全体の流れが、ふたつ目の文から MetaOCaml という特定の多段階プログラミング言語の話をして、それから応用と問題点の話になりますが、ここが多段階プログラミング一般の話なのか MetaOCaml 固有の話なのかがぶれているように思います。まずは、多段階プログラミング一般の話を応用も含めてする。(他の多段階計算ができるシステムを無視しない。)それから、MetaOCaml の紹介と、その型システムの問題点の話をすべきでしょう。MetaOCaml の型システムでできることもふれておいた方がよいでしょう。}

@@

Dependent types are types which is \AI{are} dependent on values.
We can use dependent types for securer programming. \AI{Cite Xi--Pfenning, for example?}
For example, we can realize vectors with their sizes with them.
\begin{verbatim}
        Vector :: Int -> *
\end{verbatim}
\verb|Vector| is a type constructor which takes the length of vectors, that is, 
\verb|Vector 3| is a type for vectors whose lengths are 3.
If \verb|vadd| function has the type of \verb|Vector n -> Vector n -> Vector n| for any positive number \verb|n|,
we can confirm the two arguments of \verb|vadd| function has the same length.
\red{secure programming? finer grained typing? とか}

\subsection{Multi-stage Programming with Dependent Types}
\red{このsubsectionは短いので上とくっつけてもいいかもしれない}
\AI{くっつけましょう。Section 1 で指摘した問題を依存型で解決するというのは自然なアイデアだし。One natural idea to address this problem is the introduction of dependent types. みたいなこといって、依存型の導入をする。先行研究として Brady--Hammond にもふれる。}
% \subsubsection{「生成した行列計算のコードは特定のサイズに特化しており、他のサイズでは使えない。
% このため、間違って組み合わせると容易にエラーを引き起こす。」問題を依存型で解決する。}

@@

In this paper, we design a new multi-stage calculus \LMD by 
merging dependent types into existing multi-stage calculus \LTP\cite{Hanada2014}.
\LMD is multi-stage calculus which contains dependent types.
\LMD makes multi-stage programming more safer \AI{remove ``more''} with the power of dependent types.

\AI{Give a summary of our technical contribution as itemization.}

@@

The organization of this paper is the following.
Section 2 gives an informal overview of \LMD.
Section 3 defines \LMD and
Section 4 shows the properties of \LMD including Unique Decomposition \AI{There are more important properties that should be mentioned!  In fact, the properties should be mentioned earlier.}.
Section 5 discusses future works.
Finally, Section 6 discusses related works.
\AI{Use $\backslash$label and $\backslash$ref!}
\AI{``Related work'' and ``future work'' (they are uncountable).}
\AI{Related work is discussed in Section 5?  Section 6 is the section for conclusion, which includes future work.}
